Principal Component Analysis of Population Genetic Data

### 2019-10-01 paper-bedpca.tex
add opti OADP

@@

There are 4 steps to project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data:
\begin{enumerate}
\item matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles;
\item computing PCA of the reference dataset using the matched variants only;
\item projecting computed PCs to the target data;
\item adjusting projected PCs for shrinkage.
\end{enumerate}
We implement these steps in function \texttt{bed\_projectPCA} of package bigsnpr.

For step 4, we implement a shrinkage-bias adjustment method that combines the two methods proposed by \cite{zhang2019fast}. First, for only a subset of target individuals, we compute the Online Augmentation, Decomposition, and Procrustes (OADP) transformation (using $K''=K'=K$ [REMOVE?]). For these individuals, we regress the new PC scores resulting from OADP by the projected PC scores without adjustement; this gives us shrinkage-adjustment factors that we use to correct projected PC scores for all target individuals.

@@

@@

>>>>

\subsection{Subsetted version of 1000G data [RENAME?]}

@@

To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr:
\begin{enumerate}
\item matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles;
\item computing PCA of the reference dataset using the matched variants only;
\item projecting computed PCs to the target data using an optimised implementation of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast} described in Supplementary Materials.
\end{enumerate}

%For step 4, we implement a shrinkage-bias adjustment method that combines the two methods proposed by \cite{zhang2019fast}. First, for only a subset of target individuals, we compute the Online Augmentation, Decomposition, and Procrustes (OADP) transformation (using $K''=K'=K$ [REMOVE?]). For these individuals, we regress the new PC scores resulting from OADP by the projected PC scores without adjustement; this gives us shrinkage-adjustment factors that we use to correct projected PC scores for all target individuals.

@@

\noindent[DISCUSS ABOUT $K''=K'=K$? FASTER, EASIER, DO NOT WANT MORE DIM?]

@@

\subsection*{Optimised OADP transformation}

We implement an optimised version of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation when using $K''=K'=K$ \cite[]{zhang2019fast}.
We assume that the $K$-partial Singular Value Decomposition (SVD) of the reference matrix $X$ (of size $n \times p$) has been computed as $U \Delta V^T$. There are several steps to perform OADP transformation for each sample $y$ (of size $1 \times p$) of target matrix $Y$ (of size $m \times p$):
\begin{enumerate}
\item Calculate $l = y \cdot V$ (of size $1 \times K$), where $V$ are the K PC loadings. And $g = y \cdot h^T$ (of size $1 \times 1$), where $h = (y - l \cdot V^T) / ||y - l \cdot V^T||_2$. Actually, $||y - l \cdot V^T||_2^2 = y \cdot y^T - 2 \cdot y \cdot V \cdot l^T  + l \cdot V^T \cdot V \cdot l^T = y \cdot y^T - l \cdot l^T$ and $y \cdot (y - l \cdot V^T)^T = y \cdot y^T - y \cdot V \cdot l^T = y \cdot y^T - l \cdot l^T$. Then $g = \sqrt{y \cdot y^T - l \cdot l^T}$.
\item Calculate $Q^T Q$ where $$Q = \begin{bmatrix} \Delta & l^T \\ 0 & g \end{bmatrix}$$ so that $$Q^TQ = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & g^2 + l \cdot l^T \end{bmatrix} = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & y \cdot y^T \end{bmatrix}.$$ Note that we do not actually need to compute $g$, and that we can update the last row and column of $Q^T Q$ instead of computing it from an updated version of $Q$.
\item Get the eigen decomposition $Q^T Q = V' {\Delta'}^2 {V'}^T$ (truncated to $K$ components out of the $K+1$).
\item Calculate $$\begin{bmatrix} \widetilde{U} \\ \widetilde{u}  \end{bmatrix} = \begin{bmatrix} U & 0 \\ 0 & 1 \end{bmatrix} V' = \begin{bmatrix} U V'[1{:}K,~] \\ V'[K{+}1,~] \end{bmatrix} $$
\item Find the Procrustes transformation from $\widetilde{U}$ to $U$. As $\widetilde{U}$ and $U$ have both their columns centered already, the Procrustes transformation $\rho ~\widetilde{U} A$, where $\rho$ is a scaling coefficient and $A$ is an orthonomal projection matrix that minimise the Frobenius norm of $(\rho \widetilde{U} A - U)$, is given by $A = V'' {U''}^T$ and $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}(\widetilde{U}^T \widetilde{U})}$ where $U'' \Delta'' {V''}^T$ is the full SVD of $U^T \widetilde{U} = V'[1{:}K,~]$ \cite[]{wang2015improved}. Then, $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}(\widetilde{U}^T \widetilde{U})} = \frac{\text{trace}(\Delta'')}{\text{trace}({V'[1{:}K,~]}^T V'[1{:}K,~])} = \frac{\text{trace}(\Delta'')}{\text{trace}(\Delta''^2)}$. 
\item Apply the previous transformation to $\widetilde{u}$ to get the projection of $y$ in the reference PCA space (i.e.\ $\rho \widetilde{u} A$),.
\item Finally, multipy by $\Delta$.
\end{enumerate}

[CAN COMPUTE SVD OF $V'[1{:}K,~]$ FROM LOW-RANK UPDATE?]
